---
title: "Lab 02: Introduction to the R language (1)"
author: "Seoncheol Park"
format: 
  pdf:
    logo: images/hanyang.png
    theme: [simple, css/hanyang.scss]
    geometry:
      - left=.2in
      - right=.2in
      - top=.8in
      - bottom=1.2in
    linkcolor: Cerulean
    citecolor: Fuchsia
    urlcolor: RedOrange
    toccolor: Black
    include-in-header:
      text: |
        \usepackage{amsthm, amsmath, kotex, bm, indentfirst, lmodern, graphicx, hyperref, multicol, footmisc, babel, comment, makeidx, amsfonts, amssymb}
        \usepackage{xeCJK}
        \xeCJKsetup{%
          CJKspace=true,% % true이면 띄어쓰기 사용. 중국어, 일어는 필요 없을수도
          % CJKmath=true,%  % true면 math environment 안에서 CJK 글자 사용
          CJKecglue={}%   % Western과 CJK 사이의 공백 지정: {}로 간격을 없앰
        }
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \setlength\headheight{28pt}
        \fancyhead[RO,LE]{\includegraphics[width=5cm]{images/hanyang.png}}
        \fancyfoot[LE,RO]{}
        \setCJKmainfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumMyeongjo}
        \setCJKsansfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumGothic}
        \setCJKmonofont{NanumGothicCoding}
    number-sections: true
    number-depth: 3
    mainfont: HelveticaNeue
    sansfont: HelveticaNeue
    monofont: Fira Mono
    monofontoptions: 
      - Scale=1
      - Color=orange
    CJKmainfont: NanumGothic
    classoption: twoside
pdf-engine: xelatex
---

## Instructions

<!--
author: "Seoncheol Park"
format: 
  pdf:
    logo: images/hanyang.png
    theme: [simple, css/hanyang.scss]
    geometry:
      - left=.2in
      - right=.2in
      - top=.6in
      - bottom=1.2in
    linkcolor: Cerulean
    citecolor: Fuchsia
    urlcolor: RedOrange
    toccolor: Black
    include-in-header:
      text: |
        \usepackage{amsthm, amsmath, kotex, bm, indentfirst, lmodern, graphicx, hyperref, multicol, footmisc, babel, comment, makeidx, amsfonts, amssymb}
        \usepackage{xeCJK}
        \xeCJKsetup{%
          CJKspace=true,% % true이면 띄어쓰기 사용. 중국어, 일어는 필요 없을수도
          % CJKmath=true,%  % true면 math environment 안에서 CJK 글자 사용
          CJKecglue={}%   % Western과 CJK 사이의 공백 지정: {}로 간격을 없앰
        }
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \setlength\headheight{28pt}
        \fancyhead[L]{\includegraphics[width=5cm]{images/hanyang.png}}
        \fancyfoot[LE,RO]{}
        \setCJKmainfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumMyeongjo}
        \setCJKsansfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumGothic}
        \setCJKmonofont{NanumGothicCoding}
    number-sections: true
    number-depth: 3
    mainfont: HelveticaNeue
    sansfont: HelveticaNeue
    monofont: Fira Mono
    monofontoptions: 
      - Scale=1
      - Color=orange
    CJKmainfont: NanumGothic
pdf-engine: xelatex
---
-->

- Students should solve excercises provided in the in-class exercise session on every Friday.

- Students should write (a single) R code about answers of excercises, submit it on course website.
  + [for 11628 (Fri 10:30~12:00)](https://learning.hanyang.ac.kr/courses/133489)
  + [for 13300 (Fri 09:00~10:30)](https://learning.hanyang.ac.kr/courses/133490)
  
- Please **check** whether your answers are uploaded well.

- We will **not** give any scores for all **late submission**. Please keep the time.

- You may leave early after submitting your answers on the course website.

\newpage

\vspace{5cm}

## Exercise A: R as a calculator

<!--
Textbook 2.1
-->

1. Write R code to calculate the area of a circle with radius $7$ cm.

\vspace{3cm}

2. Write R code to calculate the respective areas of the circles having radii $3, 4, \ldots, 100$.

\vspace{3cm}

3. Write a R function to calculate the area of a circle with radius $r$, where $r>0$.

\newpage

\vspace{1cm}

## Exercise B: Vectors in R

<!--
Textbook 2.3
-->

1. Write a R code to calculate the sum $\sum_{j=0}^n r^j$, where $r$ has been assigned the value $1.06, 1.08$, for $n=10,20,30$ and $40$.

\vspace{3cm}

2. Write a R code to calculate the sum $\sum_{j=0}^n r^j$, where $r$ has been assigned the value $1.08$, using the formula $(1-r^{n+1})/(1-r)$, for $n=10, 20, 30$, and $40$.

\newpage

\vspace{1cm}

## Exercise C: `rep()` and `seq()`

<!--
Textbook 2.3
-->

1. Using `rep()` and `seq()` as needed, create the vector

```{r, echo=F}
rep(seq(0,4), each=5)

```

\vspace{3cm}

2. Using `rep()` and `seq()` as needed, create the vector

```{r, echo=F}
rep(seq(1,5), 5)

```

\vspace{3cm}

3. Using `rep()`, `seq()`, and `+` as needed, create the vector

```{r, echo=F, highlight=TRUE}
rep( seq(1,5), times = 5) + rep( 0:4, each = 5)

```

\newpage

\vspace{1cm}

## Exercise D: Assigning variables

- First, note that names of variables in R may contain lowercase or capital letters, numbers, `.`, and `_`. The name must begin with a letter or `.`, and if it begins with `.`, the next character cannot be a number.

- You can use `;` to seperate multiple assignments on the same line. For example,

```{r}
#addition and subtraction
2 + 2; 2 - 2

```

We want to make some names variables in R to covert between time units. For example, you can define `min.` using `sec.` variable.

```{r}
sec. <- 1; min. <- 60*sec.

```

\vspace{1cm}

1. Write a R code to define `hr.`, `day.` `week.` using `min.`, `hr.` and `day`, respectively.

```{r, error=T, eval=F}
#define `hr.` using `min.`
hr. <- 
#define `day.` using `hr.`
day. <-
#define `week.` using `day.`
week. <-
  
```

\vspace{3cm}

2. We can also define `yr.` and `century.` variable. What is the answer of `3*century./sec.` in R?

```{r, error=T, eval=F}
yr. = 365.25*day.; century. = 100*yr.
#what is the answer?
3*century./sec.

```