---
title: "Lab 03: Introduction to the R language (2)"
author: "Seoncheol Park"
format: 
  pdf:
    logo: images/hanyang.png
    theme: [simple, css/hanyang.scss]
    geometry:
      - left=.2in
      - right=.2in
      - top=.8in
      - bottom=1.2in
    linkcolor: Cerulean
    citecolor: Fuchsia
    urlcolor: RedOrange
    toccolor: Black
    include-in-header:
      text: |
        \usepackage{amsthm, amsmath, kotex, bm, indentfirst, lmodern, graphicx, hyperref, multicol, footmisc, babel, comment, makeidx, amsfonts, amssymb}
        \usepackage{xeCJK}
        \xeCJKsetup{%
          CJKspace=true,% % true이면 띄어쓰기 사용. 중국어, 일어는 필요 없을수도
          % CJKmath=true,%  % true면 math environment 안에서 CJK 글자 사용
          CJKecglue={}%   % Western과 CJK 사이의 공백 지정: {}로 간격을 없앰
        }
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \setlength\headheight{28pt}
        \fancyhead[RO,LE]{\includegraphics[width=5cm]{images/hanyang.png}}
        \fancyfoot[LE,RO]{}
        \setCJKmainfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumMyeongjo}
        \setCJKsansfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumGothic}
        \setCJKmonofont{NanumGothicCoding}
    number-sections: false
    number-depth: 3
    mainfont: HelveticaNeue
    sansfont: HelveticaNeue
    monofont: Fira Mono
    monofontoptions: 
      - Scale=1
      - Color=orange
    highlight: monochrome
    CJKmainfont: NanumGothic
    classoption: twoside
pdf-engine: xelatex
---

## Instructions

<!--
author: "Seoncheol Park"
format: 
  pdf:
    logo: images/hanyang.png
    theme: [simple, css/hanyang.scss]
    geometry:
      - left=.2in
      - right=.2in
      - top=.6in
      - bottom=1.2in
    linkcolor: Cerulean
    citecolor: Fuchsia
    urlcolor: RedOrange
    toccolor: Black
    include-in-header:
      text: |
        \usepackage{amsthm, amsmath, kotex, bm, indentfirst, lmodern, graphicx, hyperref, multicol, footmisc, babel, comment, makeidx, amsfonts, amssymb}
        \usepackage{xeCJK}
        \xeCJKsetup{%
          CJKspace=true,% % true이면 띄어쓰기 사용. 중국어, 일어는 필요 없을수도
          % CJKmath=true,%  % true면 math environment 안에서 CJK 글자 사용
          CJKecglue={}%   % Western과 CJK 사이의 공백 지정: {}로 간격을 없앰
        }
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \setlength\headheight{28pt}
        \fancyhead[L]{\includegraphics[width=5cm]{images/hanyang.png}}
        \fancyfoot[LE,RO]{}
        \setCJKmainfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumMyeongjo}
        \setCJKsansfont[Ligatures=TeX,BoldFont={* ExtraBold},AutoFakeSlant]{NanumGothic}
        \setCJKmonofont{NanumGothicCoding}
    number-sections: true
    number-depth: 3
    mainfont: HelveticaNeue
    sansfont: HelveticaNeue
    monofont: Fira Mono
    monofontoptions: 
      - Scale=1
      - Color=orange
    CJKmainfont: NanumGothic
pdf-engine: xelatex
---
-->

- Students should solve exercises provided in the in-class exercise session every Friday.

- Students should write (a single) R code about answers to exercises and submit it on the course website.
  + [for 11628 (Fri 10:30~12:00)](https://learning.hanyang.ac.kr/courses/133489)
  + [for 13300 (Fri 09:00~10:30)](https://learning.hanyang.ac.kr/courses/133490)
  
- If you have any questions, please raise your hand. The professor and TA will help you.
  
- We will **not** give any scores for all **late submissions**. Please keep the time.

- You may leave early after submitting your answers on the course website.

- Before leaving the classroom, please **check** whether your answers are uploaded well.

\newpage

\vspace{5cm}

## Exercise A: Solar radiation observations

<!--
Exercise 2.7.1 of the textbook
-->

The following are a sample of observations on incoming solar radiation at a greenhouse.

`11.1   10.6    6.3   8.8   10.7    11.2    8.9   12.2`

```{r, echo=F,eval=F}
solar <- c(11.1, 10.6, 6.3, 8.8, 10.7, 11.2, 8.9, 12.2)
solar

```

**Exercise A-1**. Assign the data to an object called `solar.radiation`.

\vspace{3cm}

**Exercise A-2**. Find the mean, median, and variance of `solar.radiation`.

\vspace{3cm}

**Exercise A-3**. Add 10 to each observation of `solar.radiation`, and assign the result to `sr10`. Find the mean, median, range and variance of `sr10`. Which statistics change, and by how much?

\vspace{3cm}

**Exercise A-4**. Multiply each observation by `-2`, and assign the result to `srm2`. Find the mean, median, and variance of `srm2`. How do the statistics change now?

\vspace{3cm}

**Exercise A-5**. Plot a histogram of th `solar.radiation`, `sr10`, and `srm2`.

\newpage

## Exercise B: Date and times in R using `Date`, `POSIXct` and `POSIXlt`

<!--
https://rdpeng.github.io/Biostat776/lecture-working-with-dates-and-times.html

-->

<!--
Numerical analysis in science and engineering
35쪽
system.time
`POSIXlt`, `POSIXct`
?DateTimeClasses
?difftime
-->

<!--
https://shanghai.hosting.nyu.edu/data/r/dates-and-times.html
-->

- Note that R has numerous functions to manipulate times and dates. There are three basic date and time classes: `Date`, `POSIXct` and `POSIXlt`.    
  + Class `Date` handles dates **without times**.
  + `POSIXct` stores date and time in seconds with the number of seconds beginning at 1 January 1970. Negative numbers are used to store dates prior to 1970. Thus, the `POSIXct` format stores each date and time a single value in units of seconds.
  + `POSIXlt` stores dates and times as a list of components: second, minute, hour, day, month, year, time zone etc.

- **Data Tip:** The `unclass` method in R allows you to view how a particular R object is stored.

- Codes for time zones (`tz`) can be also found in this [time zone table](https://en.wikipedia.org/wiki/List_of_tz_database_time_zones).


<!--
Sys.time()

-->

```{r, message=F, warning=F}
date_1970 <- "1970-01-02 09:00:01" #Seoul: UCT + 9

datetime1 <- as.POSIXct(date_1970, tz="Asia/Seoul")
datetime1
#unclass(datetime1) #POSIXct -> numeric

#datetime2 <- as.POSIXlt(date_1970, tz="Asia/Seoul") #try it!
#datetime2
#unclass(datetime2) #POSIXct -> list

#datetime3 <- as.Date(date_1970, tz="Asia/Seoul") #try it!
#datetime3  #try it!

```
### `as.Date()`, `as.POSIXct()`, `as.POSIXlt()`

- `as.Date()`, `as.POSIXct()` and `as.POSIXlt()` convert dates and times in character forms to classes of dates and times.

- `as.Date()`, `as.POSIXct()` and `as.POSIXlt()` accept various input formats.

The default input formats are 

1. **year-month-day hour:minutes:seconds** or 

2. **year/month/day hour:minutes:seconds**.

```{r}
as.Date("2019/01/14 14:17:30")
#as.POSIXct("2019/01/14 14:17:30") #try it!
#as.POSIXlt("2019/01/14 14:17:30") #try it!

```

- If the input format is not standard, we need to set the `format` argument to map the displayed format.
  + `%b` abbreviated month name
  + `%m` month as decimal number (01–12)
  + `%c` date and time
  + `%d` day of the month as decimal number (01–31)
  + `%e` day of the month as decimal number (1–31)
  + `%H` hours as decimal number (00–23); strings such as 24:00:00 are accepted for input
  + `%I` hours as decimal number (01–12)
  + `%M` minute as decimal number (00–59)
  + `%S` second as integer (00–61)
  + `%OS` seconds including fractional seconds
  + `%Y` year with century
  + `%y` year without century (00–99)

- The full list of allowed formats can be found by `?strptime()`.

```{r}
as.Date("14jan2019 14:17:30", "%d%b%Y")
#as.POSIXct("14jan2019 14:17:30", format = "%d%b%Y %H:%M:%S")  #try it!
#as.POSIXlt("14jan2019 14:17:30", format = "%d%b%Y %H:%M:%S")  #try it!
#as.Date("14/01/2019T14:17:30", "%d/%m/%Y")  #try it!
#as.POSIXct("14/01/2019T14:17:30", format = "%d/%m/%YT%H:%M:%S", tz = "GMT")  #try it!
#as.POSIXlt("14/01/2019T14:17:30", format = "%d/%m/%YT%H:%M:%S", tz = "GMT")  #try it!

```

\newpage

**Exercise B-1**. Use all of `as.Date()`, `as.POSIXct()`, `as.POSIXlt()` functions to parse each of the following dates:

```{r}
d1 <- "06-Jun-2017 15:14:46"
d2 <- "12/30/14T19:23"

```

\vspace{3cm}

Your answers should be `Date`, `POSIXct`, and `POSIXlt` object. If you parse correctly, answers in the Console Pane may be

```{r, echo=F, message=F}
as.Date(d1, "%d-%b-%Y %H:%M:%S", tz="Asia/Seoul")
as.POSIXct(d1, format = "%d-%b-%Y %H:%M:%S", tz="Asia/Seoul") 
as.POSIXlt(d1, format = "%d-%b-%Y %H:%M:%S", tz="Asia/Seoul") 
as.Date(d2, "%m/%d/%yT%H:%M", tz="Asia/Seoul")
as.POSIXct(d2, format = "%m/%d/%yT%H:%M", tz="Asia/Seoul")
as.POSIXlt(d2, format = "%m/%d/%yT%H:%M", tz="Asia/Seoul")

```

<!--
### `srptime()`

- `strptime()` converts characters to `POSIXlt` date-time objects.

```{r}
strptime("14jan2019 14:17:30", "%d%b%Y %H:%M:%S")

strptime("14/01/2019T14:17:30", "%d/%m/%YT%H:%M:%S")

```
-->

\newpage

## Exercise C: Date and times in R using `lubridate` package

- `lubridate` provides more intuitive ways to convert characters to dates and times.

- You can install and call all functions in the `lubridate` package via

```{r, eval=F}
install.packages("lubridate")
library(lubridate)

```

```{r, echo=F, message=F, warning=F}
library(lubridate)

```

- `ymd()`, `ydm()`, `mdy()`, `myd()`, `dmy()`, `dym()`, `yq()` parses dates with year, month, and day components.

```{r}
ymd("2019/01/14")
#ydm("2019-14-01") #try it!
#dmy("14jan2019") #try it!

```

- `hm()`, `ms()`, `hms()` parses periods with hour, minute, and second components.

```{r}
hms("14:17:30")
#hm("14:17") #try it!
#ms("17:30") #try it!

```

<!--
strftime(): skip
format.POSIXlt의 wrapper라고 함
-->

- `ymd_hms()`, `ymd_hm()`, `ymd_h()`, `dmy_hms()`, `dmy_hm()`, `dmy_h()`, `mdy_hms()`, `mdy_hm()`, `mdy_h()`, `ydm_hms()`, `ydm_hm()`, `ydm_h()` parses date-times with year, month, and day, hour, minute, and second components.

```{r}
ymd_hms("2019/01/14 14:17:30")
#dmy_hms("14jan2019 14:17:30") #try it!
#dmy_hms("14/01/2019T14:17:30") #try it!

```

\newpage

**Exercise C-1**. Use the appropriate `lubridate` function to parse each of the following dates:

```{r}
e1 <- "January 1, 2023 15:14:46"
e2 <- c("August 19 (2005)", "July 1 (2005)")

```

\vspace{3cm}

If you parse correctly, answers in the Console Pane may be

```{r, echo=F}
mdy_hms(e1, tz="Asia/Seoul")
mdy(e2, tz="Asia/Seoul")

```

<!--
## Creating dates and times

### creating sequences

- We can use `seq()` to create date-time sequences.

```{r}
seq(from = as.Date("2019/01/14"), to = as.Date("2020/01/14"), by = "month")
seq(from = as.Date("2019/01/14"), to = as.Date("2019/07/14"), by = "week")

```

### combining date and time components

- We can also use `paste()` to concatenate characters and then convert them to date and time objects.

```{r}
date <- c("14jan2019", "14feb2019", "14mar2019")
time <- c("14:17:30", "15:17:30", "16:17:30")
datetime <- paste(date, time)
datetime <- as.POSIXct(datetime, format = "%d%b%Y %H:%M:%S")
datetime

```

- `make_datetime()` and `make_date()` from package lubridate create date-time objects from numeric components.

```{r}
make_datetime(year = 2019, month = 1, day = 14, hour = 14, min = 17, sec = 30)

```

## Extracting date and time components

### obtaining POSIXlt elements

- Earlier we’ve said that `POSIXlt` objects store dates and times components of second, minute, hour, day, month, year, time zone etc. as a list. To access the components, we can use the `$` operator to subset the elements we need.

```{r}
now <- Sys.time()
now <- as.POSIXlt(now)
now
now$year
now$wday
now$zone

```
-->

\newpage

## Exercise D: Calculations with dates and times

### `difftime()`

```{r}
#strptime: character -> POSIXlt
x <- strptime("2019-01-14 14:17:30", "%Y-%m-%d %H:%M:%S")
y <- strptime("2018-12-14 18:10:12", "%Y-%m-%d %H:%M:%S")
x - y

```

- The base R function `difftime()` calculates a difference of two date-time objects and returns a difftime object.

```{r}
difftime(x, y)
difftime(x, y, units = "hours") 
#difftime(x, y, units = "mins") #try it!
#difftime(x, y, units = "secs") #try it!
#difftime(x, y, units = "days") #try it!
#difftime(x, y, units = "weeks") #try it!

```

- `difftime` objects can be converted to numeric objects with `as.numeric()`.

```{r}
z <- difftime(x, y)
as.numeric(z, unit = "hours")
#as.numeric(z, unit = "mins") #try it!

```

<!--
## rounding dates and times

- Functions `round_date()`, `floor_date()` and `ceiling_date()` from `lubridate()` round the date-time objects to a nearby unit of time.

- `round_date()` takes a date-time object and rounds it to the nearest value of the specified time unit. It rounds up the object when the date-time value is exactly halfway between two consecutive units.

```{r}
datetime <- ymd_hms("2019/01/14 14:17:30.59")

round_date(datetime)

```

- `floor_date()` takes a date-time object and rounds it down to the nearest boundary of the specified time unit.

```{r}
floor_date(datetime)

```

- `ceiling_date()` takes a date-time object and rounds it up to the nearest boundary of the specified time unit.

```{r}
ceiling_date(datetime)

```

- We can set units by which to round date-time. Valid units are second, `minute`, `hour`, `day`, `week`, `month`, `bimonth`, `quarter`, `season`, `halfyear` and `year`.

```{r}
round_date(datetime, "sec")
round_date(datetime, "min")
round_date(datetime, "15 mins")
round_date(datetime, "hours")
round_date(datetime, "day")

```

## Working with time zones

- Time zones are stored as character strings of date-time objects in R. The time zone is an attribute that only controls printing.

- R relies on the user-s OS to interpret time zone names. We can get the complete list of all time zone names with `OlsonNames()`, a data base originally compiled by Arthur Olson. These names take the form `Country/City`.

- Codes for time zones can be also found in this [time zone table](https://en.wikipedia.org/wiki/List_of_tz_database_time_zones).

```{r}
head(OlsonNames())

```

- `POSIXct` and `POSIXlt` classes contain the time zone attribute.

```{r}
as.POSIXct("2019-01-14 14:17:30", format = "%Y-%m-%d %H:%M:%S", tz = "GMT")
as.POSIXlt("2019-01-14 14:17:30", format = "%Y-%m-%d %H:%M:%S", tz = "GMT")

```

- `with_tz()` and `force_tz()` from lubridate provide ways to change time zones.

- `with_tz()` displays the date-time in a different time zone, while the actual time has not been changed.

```{r}
x <- strptime("2019-01-14 14:17:30", "%Y-%m-%d %H:%M:%S")
x
with_tz(x, tzone = "America/New_York")

```

- `force_tz()` sets the date-time to a different time zone.

```{r}
x
force_tz(x, tzone = "America/New_York")

```

-->

\newpage

**Exercise D-1**. Write a function `dayspassed` that given a birthday as a `Date` object, returns how many days are passed from the given birthday. You can use `today()` function to get the current day and time.

```{r}
dayspassed <- function(aday) {
  #assume that aday is always a `Date` object
  
  #fill in the blank
}

```

\vspace{3cm}

- If you make the function `dayspassed` correctly, answers in the Console Pane may be

```{r, echo=F}
dayspassed <- function(aday) {
  #assume that aday is always a `Date` object
  
  #fill in the blank
  difftime(today(), aday)
}
```

```{r}
his_birthday <- as.Date("2019/01/14")
dayspassed(his_birthday)

```
